Notes: New gaugings in 3d from three-form

### 2023-06-05 Draft_3dH3.tex
Merging with Emanuel updates.

%  PDF specials
%\newif\ifpdf
\usepackage{ifpdf}
\ifx\pdfoutput\undefined
\pdffalse
\usepackage{cite}
\else
\pdfoutput=1
\pdftrue
%\usepackage[pdftex]{hyperref}
\pdfcompresslevel=9
\fi
\DeclareFontFamily{OMS}{rsfs}{\skewchar\font'60}
\DeclareFontShape{OMS}{rsfs}{m}{n}{<-5>rsfs5 <5-7>rsfs7 <7->rsfs10 }{}
\DeclareSymbolFont{rsfs}{OMS}{rsfs}{m}{n}
\DeclareSymbolFontAlphabet{\Scr}{rsfs}
\newcommand{\RRB}[2]{\makebox[#1]{$\left.\rule{0pt}{#2}\right)$}}
\newcommand{\LRB}[2]{\makebox[#1]{$\left(\rule{0pt}{#2}\right.$}}
\newcommand{\RSB}[2]{\makebox[#1]{$\left.\rule{0pt}{#2}\right]$}}
\newcommand{\LSB}[2]{\makebox[#1]{$\left[\rule{0pt}{#2}\right.$}}
\newcommand{\gr}[1]{\begin{greek}#1\end{greek}}
\setlength{\parskip}{0pt} \setlength{\parindent}{0.5cm}
\setcounter{footnote}{0}
\renewcommand{\theequation}{\thesection.\arabic{equation}}
\numberwithin{equation}{section}
\def\be{\begin{equation}}
	\def\ee{\end{equation}}
\def\ba{\begin{array}}
	\def\ea{\end{array}}
\newcommand{\ho}[1]{$\, ^{#1}$}
\newcommand{\hoch}[1]{$\, ^{#1}$}
\newcommand{\bea}{\begin{eqnarray}}
	\newcommand{\eea}{\end{eqnarray}}
\newcommand{\ds}{\displaystyle}
\newcommand{\BNA}{\makebox[0.8em]{\tiny$b\neq a$}}
\newcommand{\notEq}[2]{\makebox[0.8em]{\tiny$#1\neq #2$}}
\newcommand{\ph}{\alpha}
\newcommand{\scal}{z}
\newcommand{\bscal}{\bar\scal}
\newcommand{\scalC}{\pi}
%%%%%%%%%%%%%%%%% formatting
\textwidth 165mm \textheight 220mm \topmargin 0pt \oddsidemargin 2mm
%%% macros
\newcommand{\rmg}[1]{\mbox{{\small \textgreek{#1}}}}
\newcommand{\rmgg}[1]{\mathrm{#1}}
\newcommand{\ft}[2]{{\textstyle\frac{#1}{#2}}}

@@

\EM{To do:
\begin{itemize}
	\item Think about title - mention new AdS$_3$ vacua? Lifting AdS$_3$ vacua?
	\item Check the $\SL{n}$ equations
	\item Move 3-d gauged SUGRA to somewhere else??
	\item Construct 10-parameter uplift
	\item Complex geometry of moduli space?
	\item Compare with classification known in literature, e.g. in IIA (and IIB -- Jerome's story)
\end{itemize}
}

@@

Nonetheless, there are half-maximal, i.e. ${\cal N}=(4,4)$, supersymmetric AdS$_3$ vacua of string theory that are extremely intriguing. These are the AdS$_3 \times S^3 \times S^3 \times S^1$ and AdS$_3 \times S^3 \times T^4$ of IIB string theory (realising the ``small'' and ``large'' ${\cal N}=(4,4)$ superconformal symmetries), and, unlike in higher dimensions, can be supported by pure NS-NS flux. Not only does this mean that there are also heterotic versions of these vacua (preserving ${\cal N}=(4,0)$ supersymmetry), but also that they can be readily studied via the string worldsheet CFT \cite{Maldacena:2000hw,Eberhardt:2018ouy,Eberhardt:2019niq,Eberhardt:2019ywk}. Moreover, the string sigma model in these vacua is also integrable \cite{}. Finally, the AdS$_3$/CFT$_2$ correspondence has seen much progress recently, with many new holographic duals being established \cite{}.

@@

Another technical result of our work is the derivation of the potential of 3-dimensional gauged supergravity in terms of the embedding tensor, which was previously only expressed in terms of fermion shift matrices. As a by-product, we find that a constraint that must be obeyed by the embedding tensor of any 3-dimensional gauged supergravity that can be uplifted to 10-/11-dimensional supergravity. As we show, this allows us to prove a higher-dimensional origin of some 3-dimensional gaugings.

@@

Thus, for a given twist matrix $U_{M}{}^{\fl{M}}$ satisfying the consistency condition~\eqref{eq:GenLeibniz}, the action~\eqref{eq:ExFTaction} reduces to three-dimensional ${\cal N}=16$ gauged supergravity. The constant tensor $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ plays the role of the embedding tensor of the three-dimensional gauged supergravity, as is clear from its appearance in \eqref{eq:3dCovDeriv} in the gauge-covariant derivative and the gauge algebra of the vector fields. From eq. \eqref{eq:GenLeibniz}, it has the following expression in terms of the twist matrix and $\rho$
\begin{equation}
	\begin{aligned}
		X_{\fl{MN}}{}^{\fl{P}}&=-\rho^{-1}\,\Gamma_{\fl{MN}}{}^{\fl{P}} + \rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{L}}\,\Gamma_{\fl{KM}}{}^{\fl{L}}-\frac{1}{60}\,\rho^{-1}\,f^{\fl{PK}}{}_{\fl{N}}\,f_{\fl{ML}}{}^{\fl{Q}}\,\Gamma_{\fl{KQ}}{}^{\fl{L}}\\
		&\quad-\frac{1}{2}\,\rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{M}}\,\Gamma_{\fl{RK}}{}^{\fl{R}}+\left(\delta_{\fl{M}}{}^{\fl{K}}\delta_{\fl{N}}{}^{\fl{P}}-\frac{1}{2}\,f_{\fl{M}}{}^{\fl{LK}}f_{\fl{NL}}{}^{\fl{P}}\right)\,\xi_{\fl{K}}\,.
	\end{aligned}
\end{equation}
Here, we defined the ${\rm E}_{8(8)}$ current $\Gamma_{\fl{MN}}{}^{\fl{P}}=U_{\fl{M}}{}^{K}U_{\fl{N}}{}^{L}\partial_{K}U^{\fl{P}}{}_{L}$ and the trombone gauging \EM{Give this in terms of $\Gamma_{\fl{M}\fl{N}}$ for later use?}
\begin{equation} \label{eq:Trombone}
	\xi_{\fl{M}} = 2\,U_{\fl{M}}{}^{N}\partial_{N}\rho^{-1}+\rho^{-1}\,\partial_{N}U_{\fl{M}}{}^{N} \,.
\end{equation}
In the following, we consider only gaugings with $\xi_{\fl{M}}=0$. The embedding tensor is then most nicely expressed once projected on the adjoint representation \CE{(Keep expression with projector on 3875?)} \EM{I think it's useful}
\begin{equation} \label{eq:symembeddingtensor}
	\begin{aligned}
		X_{\fl{MN}} = -\frac{1}{60}\,X_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}} &= 2\,\rho^{-1}\,\Gamma_{(\fl{MN})} - \rho^{-1}\,\Gamma_{\fl{P}(\fl{M}}{}^{\fl{Q}}\,f_{\fl{N})\fl{Q}}{}^{\fl{P}}\\
		&=14\,\rho^{-1}\,(\mathbb{P}_{3875})_{\fl{MN}}{}^{\fl{PQ}}\,\Gamma_{\fl{PQ}}+\frac{1}{4}\,\rho^{-1}\,\eta_{\fl{MN}}\,\Gamma_{\fl{P}}{}^{\fl{P}} \,,
	\end{aligned}
\end{equation}
with the projection of the current $\Gamma_{\fl{MN}} = \tfrac{1}{60}\,\Gamma_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}}$.

The three-dimensional action follows from inserting the Ansätze~\eqref{eq:GSSansatz} in the ExFT action~\eqref{eq:ExFTaction}. The expressions of the kinetic and Chern-Simons terms result immediately from eq.~\eqref{eq:GSSexample} \CE{(A word on the Ricci scalar?)} \EM{Yes, should briefly mention what happens}
\begin{equation}
	\begin{aligned}
		\mathscr{L}_{\rm kin} &= \frac{1}{240}\,D_{\mu}\M_{MN}D^{\mu}\M^{MN} \underset{\rm gSS}{=} \frac{\rho^{2}}{240}\,{\cal D}_{\mu}\M_{\fl{MN}}{\cal D}^{\mu}\M^{\fl{MN}}, \\
		\mathscr{L}_{\rm CS} &\underset{\rm gSS}{=} -\rho^{-1}\,\varepsilon^{\mu\nu\rho}\,X_{\fl{MN}}\,\A_{\mu}{}^{\fl{M}} \left(\partial_{\nu}\A_{\rho}{}^{\fl{N}}-\frac{1}{3}\,\llbracket\A_{\nu},\A_{\rho}\rrbracket^{\fl{N}}\right)\,.
	\end{aligned}
\end{equation}

However, the case of the potential term $\mathscr{L}_{\rm int}$, leading to a potential $V$ for the three-dimensional scalars $\M_{\fl{MN}}$, is more subtle. We expect $V$ to be quadratic in the embedding tensor, which makes the identification of $X_{\fl{MN}}$ more complicated. Moreover, the potential of three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor is unknown, with currently the potential only expressed in terms of the fermion shift matrices of the gauged supergravity. Thus, in the following, we will follow the truncation procedure carefully and thereby construct the potential of the three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor.

\section{Deriving the potential of ${\cal N}=16$ gauged supergravity} \label{s:Potential}
Here we will use the $\EE$ ExFT Lagrangian and the consistent truncation Ansatz \eqref{eq:GSSansatz} to derive the potential of three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor. Not only is this an interesting application of ExFT, relying on purely bosonic considerations and bypassing the usual construction of the gauged supergravity potential using supersymmetry\footnote{Note that the same strategy was recently used to impressively derive the potential of maximal two-dimensional gauged supergravity \cite{Bossard:2022wvi}, where fermions are extremely poorly understood. As a result, in that case, the bosonic $\En{9}$ ExFT \cite{Bossard:2018utw,Bossard:2021jix} provides the only currently accessible route for computing the gauged supergravity potential.}, but the potential is crucial for finding vacua of the three-dimensional theory and uplifting these in the later parts of this paper.

@@

Now, we first insert the generalized Sherk-Schwarz Ansätze \eqref{eq:GSSansatz} into the expression of $\mathscr{L}_{\rm int}$ \eqref{eq:ExFTLint} with the additional total derivative \eqref{eq:BoundaryTermCombination} \CE{(Need to motivate the total derivative? Mention the need for section constraint? Factor $\sqrt{\vert g\vert}$ to avoid carrying $\rho$ in $V$?)}
\begin{equation} \label{eq:Lintexpansion}
	\begin{aligned}
		\mathscr{L}_{\rm int} &+ \frac{a}{\sqrt{\vert g\vert}}\,\partial_{M}\Big(\sqrt{\vert g\vert}\,\partial_{N}\M^{MN}+\frac{4}{3}\,\M^{MN}\partial_{N}\sqrt{\vert g\vert}\Big)\\
		&\underset{\rm gSS}{=} -\frac{1}{2}\,\M^{\fl{MN}}\,\Gamma_{\fl{MK}}\Gamma_{\fl{N}}{}^{\fl{K}} -\M^{\fl{MN}}\,\Gamma_{\fl{MK}}\Gamma^{\fl{K}}{}_{\fl{N}} -\frac{1}{2}\,\Gamma_{\fl{MN}}\Gamma^{\fl{NM}} - \frac{1}{2}\,\M^{\fl{MN}}\M^{\fl{KL}}\,\Gamma_{\fl{MK}}\Gamma_{\fl{NL}} \\
		&\ \quad - \frac{1}{2}\,\M^{\fl{MN}}\M^{\fl{KL}}\,\Gamma_{\fl{MK}}\Gamma_{\fl{LN}} +\frac{3}{2}\left(a-1\right)\,\M^{\fl{MN}}\Gamma_{\fl{MN}}{}^{\fl{K}}\Gamma_{\fl{LK}}{}^{\fl{L}}-\frac{a}{2}\,\M^{\fl{MN}}\Gamma_{\fl{KM}}{}^{\fl{K}}\Gamma_{\fl{LN}}{}^{\fl{L}}\\
		&\ \quad + \left(\frac{1}{2}-a\right)\,\M^{\fl{MN}}\Gamma_{\fl{KM}}{}^{\fl{L}}\Gamma_{\fl{LN}}{}^{\fl{K}}+\,\M^{\fl{MN}}\Gamma_{\fl{MK}}{}^{\fl{L}}\Gamma_{\fl{LN}}{}^{\fl{K}}-\frac{1}{2}\,\M^{\fl{MN}}\M^{\fl{KL}}f_{\fl{QL}}{}^{\fl{P}}\Gamma_{\fl{PN}}{}^{\fl{Q}}\Gamma_{\fl{MK}}\,.
	\end{aligned}
\end{equation}

Secondly, knowing that the ${\cal N}=16$ potential must be quadratic in $X_{\fl{MN}}$, we consider the most general quadratic function in $X_{\fl{MN}}$ and develop it in \EE currents using eq.~\eqref{eq:symembeddingtensor}. \CE{(Sign convention between $V$ and ${\mathscr{L}}_{\rm int}$? $+{\mathscr{L}}_{\rm int}$ in eq.~\eqref{eq:ExFTaction} but $-V$ in eq.~\eqref{eq:ExFTactiongSS}.)} Note that in higher dimensions, there exist only two quadratic combinations of the embedding tensor
\begin{equation}
	X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{P}}{}^{\fl{N}} \gM^{\fl{MQ}} \qquad \text{and} \qquad X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{R}}{}^{\fl{S}} \gM^{\fl{MQ}} \gM^{\fl{N}\fl{R}} \gM_{\fl{P}\fl{S}} \,.
\end{equation}
However, in three dimensions, because the vector fields transform in the adjoint representation of $\EE$, we can write two additional terms
\begin{equation} \label{eq:ExtraX2Terms}
	X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MP}} \eta^{\fl{NQ}} \qquad \text{and} \qquad X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MN}} \eta^{\fl{PQ}} \,,
\end{equation}
where the second term corresponds to the square of the singlet part of the embedding tensor.

It turns out that for gaugings that arise from consistent truncations, these two terms \eqref{eq:ExtraX2Terms} can be related to the square of the trombone tensor. We can square eq.~\eqref{eq:symembeddingtensor} to derive an expression for $X_{\ov{MN}}X^{\ov{MN}}$ in terms of the current $\Gamma$. We find 
\begin{equation}\label{eq:XXwitheta}
	X_{\ov{MN}}X^{\ov{MN}} = 21 \rho^{-2}\, \Gamma_{\ov{MN}}\, \Gamma^{\ov{NM}} + 19 \rho^{-2}\left(\Gamma_{\ov M}{}^{\ov M}\right)^2\,.
\end{equation}
Similarly, squaring the trombone tensor $\xi$ gives \EM{Should we give \eqref{eq:Trombone} in terms of $\Gamma$?}, we find
\begin{equation}\label{eq:xixiwitheta}
	\xi_{\ov M}\,\xi^{\ov M}=\rho^{-2}\left(\Gamma_{\ov M}{}^{\ov M}\Gamma_{\ov N}{}^{\ov N}+\Gamma_{\ov{MN}}\Gamma^{\ov{NM}}\right) \,.
\end{equation}
We can now deduce
\begin{equation}
	X_{\ov{MN}} \,X^{\ov{MN}}=21	\xi_{\ov M} \, \xi^{\ov M} - 2 \rho^{-2}\left(\Gamma_{\ov M}{}^{\ov M}\right)^2 \,,
\end{equation}
which can be expressed entirely in terms of $X_{\fl{M}\fl{N}}$ after tracing eq.~\eqref{eq:symembeddingtensor}
\begin{equation}\label{eq:reltrX}
	X_{\ov{MN}}\,X^{\ov{MN}} = 21\, \xi_{\ov M}\, \xi^{\ov M} - \frac1{1922}\left(X_{\ov M}{}^{\ov M}\right)^2 \,. 
\end{equation}
Thus, \eqref{eq:reltrX} is a constraint on the embedding tensor of three-dimensional supergravity which must be satisfied in order for it to have an uplift to 10-/11-dimensional supergravity. Note that this constraint is not implied by the quadratic constraint. To see this, simply observe that the theory with $\En{8}$ gauging has $X_{\ov{MN}}=\eta_{\ov{MN}}$ and $\xi_{\fl{M}} = 0$, violating \eqref{eq:reltrX}.

>>>>

%  PDF specials
%\newif\ifpdf
\usepackage{ifpdf}
\ifx\pdfoutput\undefined
\pdffalse
\usepackage{cite}
\else
\pdfoutput=1
\pdftrue
%\usepackage[pdftex]{hyperref}
\pdfcompresslevel=9
\fi
\DeclareFontFamily{OMS}{rsfs}{\skewchar\font'60}
\DeclareFontShape{OMS}{rsfs}{m}{n}{<-5>rsfs5 <5-7>rsfs7 <7->rsfs10 }{}
\DeclareSymbolFont{rsfs}{OMS}{rsfs}{m}{n}
\DeclareSymbolFontAlphabet{\Scr}{rsfs}
\newcommand{\RRB}[2]{\makebox[#1]{$\left.\rule{0pt}{#2}\right)$}}
\newcommand{\LRB}[2]{\makebox[#1]{$\left(\rule{0pt}{#2}\right.$}}
\newcommand{\RSB}[2]{\makebox[#1]{$\left.\rule{0pt}{#2}\right]$}}
\newcommand{\LSB}[2]{\makebox[#1]{$\left[\rule{0pt}{#2}\right.$}}
\newcommand{\gr}[1]{\begin{greek}#1\end{greek}}
\setlength{\parskip}{0pt} \setlength{\parindent}{0.5cm}
\setcounter{footnote}{0}
\renewcommand{\theequation}{\thesection.\arabic{equation}}
\numberwithin{equation}{section}
\def\be{\begin{equation}}
	\def\ee{\end{equation}}
\def\ba{\begin{array}}
	\def\ea{\end{array}}
\newcommand{\ho}[1]{$\, ^{#1}$}
\newcommand{\hoch}[1]{$\, ^{#1}$}
\newcommand{\bea}{\begin{eqnarray}}
	\newcommand{\eea}{\end{eqnarray}}
\newcommand{\ds}{\displaystyle}
\newcommand{\BNA}{\makebox[0.8em]{\tiny$b\neq a$}}
\newcommand{\notEq}[2]{\makebox[0.8em]{\tiny$#1\neq #2$}}
\newcommand{\ph}{\alpha}
\newcommand{\scal}{z}
\newcommand{\bscal}{\bar\scal}
\newcommand{\scalC}{\pi}
%%%%%%%%%%%%%%%%% formatting
\textwidth 165mm \textheight 220mm \topmargin 0pt \oddsidemargin 2mm
%%% macros
\newcommand{\rmg}[1]{\mbox{{\small \textgreek{#1}}}}
\newcommand{\rmgg}[1]{\mathrm{#1}}
\newcommand{\ft}[2]{{\textstyle\frac{#1}{#2}}}
\newcommand{\ts}[1]{\widetilde{#1}}

@@

\EM{To do:
\begin{itemize}
	\item Think about title - mention new AdS$_3$ vacua? Lifting AdS$_3$ vacua?
	\item Check the $\SL{n}$ equations
	\item Move 3-d gauged SUGRA to somewhere else??
	\item Construct 10-parameter uplift
	\item Complex geometry of moduli space?
	\item Compare with classification known in literature, e.g. in IIA (and IIB -- Jerome's story)
	\item Email DW and MP to ask who we should cite for \eqref{eq:Twisting}
\end{itemize}
}

@@

Nonetheless, there are half-maximal, i.e. ${\cal N}=(4,4)$, supersymmetric AdS$_3$ vacua of string theory that are extremely intriguing. These are the AdS$_3 \times S^3 \times S^3 \times S^1$ and AdS$_3 \times S^3 \times T^4$ of IIB string theory (realising the ``large'' and ``small'' ${\cal N}=(4,4)$ superconformal symmetries), and, unlike in higher dimensions, can be supported by pure NS-NS flux. Not only does this mean that there are also heterotic versions of these vacua (preserving ${\cal N}=(4,0)$ supersymmetry), but also that they can be readily studied via the string worldsheet CFT \cite{Maldacena:2000hw,Eberhardt:2018ouy,Eberhardt:2019niq,Eberhardt:2019ywk}. Moreover, the string sigma model in these vacua is also integrable \cite{}. Finally, the AdS$_3$/CFT$_2$ correspondence has seen much progress recently, with many new holographic duals being established \cite{}.

@@

Another technical result of our work is the derivation of the potential of 3-dimensional gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv} in terms of the embedding tensor, which was previously only expressed in terms of fermion shift matrices. As a by-product, we find that a constraint that must be obeyed by the embedding tensor of any 3-dimensional gauged supergravity that can be uplifted to 10-/11-dimensional supergravity. As we show, this allows us to prove a higher-dimensional origin of some 3-dimensional gaugings.

@@

Thus, for a given twist matrix $U_{M}{}^{\fl{M}}$ satisfying the consistency condition~\eqref{eq:GenLeibniz}, the action~\eqref{eq:ExFTaction} reduces to three-dimensional ${\cal N}=16$ gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv}. The constant tensor $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ plays the role of the embedding tensor of the three-dimensional gauged supergravity, as is clear from its appearance in \eqref{eq:3dCovDeriv} in the gauge-covariant derivative and the gauge algebra of the vector fields. From eq. \eqref{eq:GenLeibniz}, it has the following expression in terms of the twist matrix and $\rho$
\begin{equation}
	\begin{aligned}
		X_{\fl{MN}}{}^{\fl{P}}&=-\rho^{-1}\,\Gamma_{\fl{MN}}{}^{\fl{P}} + \rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{L}}\,\Gamma_{\fl{KM}}{}^{\fl{L}}-\frac{1}{60}\,\rho^{-1}\,f^{\fl{PK}}{}_{\fl{N}}\,f_{\fl{ML}}{}^{\fl{Q}}\,\Gamma_{\fl{KQ}}{}^{\fl{L}}\\
		&\quad-\frac{1}{2}\,\rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{M}}\,\Gamma_{\fl{RK}}{}^{\fl{R}}+\left(\delta_{\fl{M}}{}^{\fl{K}}\delta_{\fl{N}}{}^{\fl{P}}-\frac{1}{2}\,f_{\fl{M}}{}^{\fl{LK}}f_{\fl{NL}}{}^{\fl{P}}\right)\,\xi_{\fl{K}}\,.
	\end{aligned}
\end{equation}
Here, we defined the ${\rm E}_{8(8)}$ current $\Gamma_{\fl{MN}}{}^{\fl{P}}=U_{\fl{M}}{}^{K}U_{\fl{N}}{}^{L}\partial_{K}U^{\fl{P}}{}_{L}$ and the trombone gauging
\begin{equation} \label{eq:Trombone}
	\begin{split}
		\xi_{\fl{M}} %&= 2\,U_{\fl{M}}{}^{N}\partial_{N}\rho^{-1}+\rho^{-1}\,\partial_{N}U_{\fl{M}}{}^{N} \\
		&= 2\,U_{\fl{M}}{}^{N}\partial_{N}\rho^{-1} - \rho^{-1}\, \Gamma_{\ov{N}\ov{M}}{}^{\ov{N}} \,.
	\end{split}
\end{equation}
The embedding tensor is most nicely expressed once projected on the adjoint representation
\begin{equation} \label{eq:symembeddingtensor}
	\begin{aligned}
		X_{\fl{MN}} = -\frac{1}{60}\,X_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}} &= - 2\,\rho^{-1}\,\Gamma_{(\fl{MN})} - \rho^{-1}\,\Gamma_{\fl{P}(\fl{M}}{}^{\fl{Q}}\,f_{\fl{N})\fl{Q}}{}^{\fl{P}} + \frac12 f_{\ov{M}\ov{N}}{}^{\ov{P}}\, \xi_{\ov{P}} \\
		&= - 14\,\rho^{-1}\,(\mathbb{P}_{3875})_{\fl{MN}}{}^{\fl{PQ}}\,\Gamma_{\fl{PQ}} - \frac{1}{4}\,\rho^{-1}\,\eta_{\fl{MN}}\,\Gamma_{\fl{P}}{}^{\fl{P}} + \frac12 f_{\ov{M}\ov{N}}{}^{\ov{P}}\, \xi_{\ov{P}} \,,
	\end{aligned}
\end{equation}
with the projection of the current $\Gamma_{\fl{MN}} = - \tfrac{1}{60}\,\Gamma_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}}$. In the following, we will focus on gaugings with $\xi_{\fl{M}}=0$.

The three-dimensional action follows from inserting the Ansätze~\eqref{eq:GSSansatz} in the ExFT action~\eqref{eq:ExFTaction}. The expressions of the kinetic and Chern-Simons terms result immediately from eq.~\eqref{eq:GSSexample}
\begin{equation}
	\begin{aligned}
		\mathscr{L}_{\rm kin} &= \frac{1}{240}\,D_{\mu}\M_{MN}D^{\mu}\M^{MN} \underset{\rm gSS}{=} \frac{\rho^{2}}{240}\,{\cal D}_{\mu}\M_{\fl{MN}}{\cal D}^{\mu}\M^{\fl{MN}}, \\
		\mathscr{L}_{\rm CS} &\underset{\rm gSS}{=} -\rho^{-1}\,\varepsilon^{\mu\nu\rho}\,X_{\fl{MN}}\,\A_{\mu}{}^{\fl{M}} \left(\partial_{\nu}\A_{\rho}{}^{\fl{N}}-\frac{1}{3}\,\llbracket\A_{\nu},\A_{\rho}\rrbracket^{\fl{N}}\right)\,.
	\end{aligned}
\end{equation}

However, the case of the potential term $\mathscr{L}_{\rm int}$, leading to a potential $V$ for the three-dimensional scalars $\M_{\fl{MN}}$, is more subtle. We expect $V$ to be quadratic in the embedding tensor, which makes the identification of $X_{\fl{MN}}$ more complicated. Moreover, the potential of three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor is unknown, with currently the potential only expressed in terms of the fermion shift matrices of the gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv}. Thus, in the following, we will follow the truncation procedure carefully and thereby construct the potential of the three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor.

\section{Deriving the potential of ${\cal N}=16$ gauged supergravity} \label{s:Potential}
Here we will use the $\EE$ ExFT Lagrangian and the consistent truncation Ansatz \eqref{eq:GSSansatz} to derive the potential of three-dimensional ${\cal N}=16$ gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv} in terms of the embedding tensor. Not only is this an interesting application of ExFT, relying on purely bosonic considerations and bypassing the usual construction of the gauged supergravity potential using supersymmetry\footnote{Note that the same strategy was recently used to impressively derive the potential of maximal two-dimensional gauged supergravity \cite{Bossard:2022wvi}, where fermions are extremely poorly understood. As a result, in that case, the bosonic $\En{9}$ ExFT \cite{Bossard:2018utw,Bossard:2021jix} provides the only currently accessible route for computing the gauged supergravity potential.}, but the potential is crucial for finding vacua of the three-dimensional theory and uplifting these in the later parts of this paper.

@@

Now, we first insert the generalized Sherk-Schwarz Ansätze \eqref{eq:GSSansatz} into the expression of $\mathscr{L}_{\rm int}$ \eqref{eq:ExFTLint} with the additional total derivative \eqref{eq:BoundaryTermCombination}
\begin{equation} \label{eq:Lintexpansion}
	\begin{aligned}
		\mathscr{L}_{\rm int} &+ \frac{a}{\sqrt{\vert g\vert}}\,\partial_{M}\Big(\sqrt{\vert g\vert}\,\partial_{N}\M^{MN}+\frac{4}{3}\,\M^{MN}\partial_{N}\sqrt{\vert g\vert}\Big)\\
		&\underset{\rm gSS}{=} -\frac{1}{2}\,\M^{\fl{MN}}\,\Gamma_{\fl{MK}}\Gamma_{\fl{N}}{}^{\fl{K}} -\M^{\fl{MN}}\,\Gamma_{\fl{MK}}\Gamma^{\fl{K}}{}_{\fl{N}} -\frac{1}{2}\,\Gamma_{\fl{MN}}\Gamma^{\fl{NM}} - \frac{1}{2}\,\M^{\fl{MN}}\M^{\fl{KL}}\,\Gamma_{\fl{MK}}\Gamma_{\fl{NL}} \\
		&\ \quad - \frac{1}{2}\,\M^{\fl{MN}}\M^{\fl{KL}}\,\Gamma_{\fl{MK}}\Gamma_{\fl{LN}} +\frac{3}{2}\left(a-1\right)\,\M^{\fl{MN}}\Gamma_{\fl{MN}}{}^{\fl{K}}\Gamma_{\fl{LK}}{}^{\fl{L}}-\frac{a}{2}\,\M^{\fl{MN}}\Gamma_{\fl{KM}}{}^{\fl{K}}\Gamma_{\fl{LN}}{}^{\fl{L}}\\
		&\ \quad + \left(\frac{1}{2}-a\right)\,\M^{\fl{MN}}\Gamma_{\fl{KM}}{}^{\fl{L}}\Gamma_{\fl{LN}}{}^{\fl{K}}+\,\M^{\fl{MN}}\Gamma_{\fl{MK}}{}^{\fl{L}}\Gamma_{\fl{LN}}{}^{\fl{K}}-\frac{1}{2}\,\M^{\fl{MN}}\M^{\fl{KL}}f_{\fl{QL}}{}^{\fl{P}}\Gamma_{\fl{PN}}{}^{\fl{Q}}\Gamma_{\fl{MK}}\,.
	\end{aligned}
\end{equation}

Secondly, knowing that the ${\cal N}=16$ potential must be quadratic in $X_{\fl{MN}}$, we consider the most general quadratic function in $X_{\fl{MN}}$ and develop it in \EE currents using eq.~\eqref{eq:symembeddingtensor}. Note that in higher dimensions, there exist only two quadratic combinations of the embedding tensor
\begin{equation}
	X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{P}}{}^{\fl{N}} \gM^{\fl{MQ}} \qquad \text{and} \qquad X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{R}}{}^{\fl{S}} \gM^{\fl{MQ}} \gM^{\fl{N}\fl{R}} \gM_{\fl{P}\fl{S}} \,.
\end{equation}
However, in three dimensions, because the vector fields transform in the adjoint representation of $\EE$, we can write two additional terms
\begin{equation} \label{eq:ExtraX2Terms}
	X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MP}} \eta^{\fl{NQ}} \qquad \text{and} \qquad X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MN}} \eta^{\fl{PQ}} \,,
\end{equation}
where the second term corresponds to the square of the singlet part of the embedding tensor.

It turns out that for gaugings that arise from consistent truncations, these two terms \eqref{eq:ExtraX2Terms} can be related to the square of the trombone tensor. We can square eq.~\eqref{eq:symembeddingtensor} to derive an expression for $X_{\ov{MN}}X^{\ov{MN}}$ in terms of the current $\Gamma$. We find \EM{check coefficient of last term}
\begin{equation}\label{eq:XXwitheta}
	X_{\ov{MN}}X^{\ov{MN}} = 21 \rho^{-2}\, \Gamma_{\ov{MN}}\, \Gamma^{\ov{NM}} + 19 \rho^{-2}\left(\Gamma_{\ov M}{}^{\ov M}\right)^2 - 15\, \xi_{\ov{M}}\, \xi^{\ov{M}} \,.
\end{equation}
Similarly, squaring the trombone tensor $\xi$ gives
\begin{equation}\label{eq:xixiwitheta}
	\xi_{\ov M}\,\xi^{\ov M}=\rho^{-2}\left(\Gamma_{\ov M}{}^{\ov M}\Gamma_{\ov N}{}^{\ov N}+\Gamma_{\ov{MN}}\Gamma^{\ov{NM}}\right) \,.
\end{equation}
We can now deduce
\begin{equation}
	X_{\ov{MN}} \,X^{\ov{MN}}= 6\, \xi_{\ov M} \, \xi^{\ov M} - 2\, \rho^{-2}\left(\Gamma_{\ov M}{}^{\ov M}\right)^2 \,,
\end{equation}
which can be expressed entirely in terms of $X_{\fl{M}\fl{N}}$ after tracing eq.~\eqref{eq:symembeddingtensor}
\begin{equation}\label{eq:reltrX}
	X_{\ov{MN}}\,X^{\ov{MN}} = 6\, \xi_{\ov M}\, \xi^{\ov M} - \frac1{1922}\left(X_{\ov M}{}^{\ov M}\right)^2 \,. 
\end{equation}
Thus, \eqref{eq:reltrX} is a constraint on the embedding tensor of three-dimensional supergravity which must be satisfied in order for it to have an uplift to 10-/11-dimensional supergravity. Note that this constraint is not implied by the quadratic constraint. To see this, simply observe that the theory with $\En{8}$ gauging has $X_{\ov{MN}}=\eta_{\ov{MN}}$ and $\xi_{\fl{M}} = 0$ \cite{Nicolai:2000sc,Nicolai:2001sv}, violating \eqref{eq:reltrX}.
